% This file is part of the spectral-variability project.
% Copyright 2016, 2017 the authors.

\documentclass[12pt, letterpaper]{article}
\usepackage{amsmath}

\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\Avast}{\project{A$\!$vast}}
\newcommand{\harps}{\project{\acronym{HARPS}}}

\newcommand{\unit}[1]{\mathrm{#1}}
\newcommand{\m}{\unit{m}}
\newcommand{\s}{\unit{s}}
\newcommand{\mps}{\m\,\s^{-1}}

\newcommand{\dd}{\mathrm{d}}
\newcommand{\given}{\,|\,}
\newcommand{\normal}{{\mathcal{N}}}
\newcommand{\com}{\mathrm{cm}}
\newcommand{\surf}{\mathrm{sf}}

\setlength{\parindent}{1.0\baselineskip}
\linespread{1.09}
\raggedbottom
\sloppy
\sloppypar
\frenchspacing

\begin{document}

\section*{The variations with time of stellar spectra}

\noindent
by MB, DWH, others

\paragraph{Abstract:}
% Context:
Stellar surfaces churn with convection and rotation, carry
time-variable star spots, and magnetically flare.
A stellar spectrum ought to be a time-dependent mixture of temperatures
and radiative processes.
Indeed, 3-dimensional models of stars predict such variability, though
it ought to be subtle.
% Aims:
In this project, we look for variations in stellar spectra for a few
individual main-sequence stars that have many epochs of high
resolution, high signal-to-noise spectroscopy taken with the
\harps\ instrument.
% Methods:
Because the variations with time are subtle, we build an extremely
precise (signal-to-noise in the thousands) smooth stellar spectrum
model using all the specral epochs simultaneously, accounting for tiny
radial-velocity changes, and significant throughput changes from epoch
to epoch.
We inspect, and take the variance of, the residuals away from this
model as a function of wavelength to find time-variable narrow
spectral features.
% Results:
We find YYY.

\section{Introduction}


\section{Method}

The most important assumptions we make in this work are the following:
\begin{itemize}\itemsep=0ex
\item There are $N$ multiple extracted spectra of the same star, taken
  at (fictitious) Solar-System barycentric times (epochs) $t_n$. Each
  of these spectra has been continuum normalized properly, and is
  represented on a grid of $M$ wavelength pixels $\lambda_m$ in the
  (fictitious) Solar-System barycentric rest frame. We assume (perhaps
  naively) that both the times $t_n$ and wavelengths $\lambda_m$ are
  very precisely known.
\item At each pixel $m$ of each spectrum $n$ there is Gaussian noise,
  drawn from a distribution with zero mean and known variance
  $\sigma^2_{ij}$.
\item At each epoch $n$, there is a radial velocity displacement
  $\Delta v_{n}$ of the star. This shift may be a combination
  of ``true'' center-of-mass radial velocity induced by stellar companions
  and/or (integrated) velocity fields in the stellar photosphere. Additionally,
  there is also some kind of spectral distortion.  That is, there is a
  composite Doppler shift and a spectral change at every epoch.
\item We expect some aspects of the spectral distortions to be sparse
  in wavelength-space.  That is, we expect only a few, isolated lines
  to participate in (say) a distortion arising from chromospheric stellar activity.
\item We expect other aspects of the spectral distortions to be low
  rank; these might be caused by systematic problems in the
  spectrograph, or in the continuum normalization procedures.
\item We expect that the center-of-mass velocity is set by some kind
  of kinematic or dynamical model (such as the two-body orbit problem)
  and any distortions to be set by some kind of stochastic process.
\end{itemize}

Our model (likelihood) is described by
\begin{eqnarray}
  %y_{ij} &=& s_j f(\lambda'_{ij}) + \mathrm{systematics}_{ij} + \mathrm{noise}_{ij}
   y_{ij} &=& s_j f(\lambda'_{ij})
  \\
  \lambda'_{ij} &=& \lambda_i\,\sqrt{\frac{1 - \beta_j}{1 + \beta_j}}
  \\
  \beta &\equiv& \frac{\Delta v}{c}
  \quad , 
\end{eqnarray}
where the $y_{ij}$ is the continuum-normalized flux in pixel $i$ of
spectrum $j$, $s_j$ is a constant scaling factor related to the observed
flux level at epoch $j$, $f()$ is a function of
wavelength describing the intrinsic stellar spectrum, and
$\lambda'_{ij}$ is the stellar rest-frame wavelength
corresponding to the barycentric wavelength $\lambda_i$ observed at
(dimensionless) velocity $\beta_j$.
%There are both (independent-ish, pixel) noise contributions and
%(covariant, low-rank) systematics contributions to the observed
%signal.
We represent or generate these functions and latent variables
with the following:
\begin{eqnarray}
  f(\lambda) &=& \sum_{m=0}^{M-1} a_m\,g_m(\lambda)
  \\
  g_m(\lambda) &=& \left\{
        \begin{array}{ll}
            0 & \quad \lambda \leq x_m -\Delta x \\
            1 - \frac{x_m - \lambda}{\Delta x} & \quad x_m - \Delta x \leq \lambda \leq x_m \\
            1 - \frac{\lambda - x_m}{\Delta x} & \quad x_m \leq \lambda \leq x_m + \Delta x \\
            0 & \quad \lambda \geq x_m + \Delta x \label{eqn:triangle}
        \end{array}
    \right.
  \quad ,
\end{eqnarray}
where the $a_m$ are expansion coefficients and the $g_m(\lambda)$
are basis functions corresponding to a triangle centered at $x_m$ with half-width 
$\Delta x$. For this implementation, we chose $x_m$ spaced evenly in $\log \lambda$, 
with $\Delta x$ equal to the spacing between each $x_m$.
%We also have both systematics and noise, which we model with covariant
%and independent Gaussians:
%\begin{eqnarray}
%  \mathrm{systematics}_{ij} &=& \sum_{t=0}^{T-1} A_{jt}\,G_t(\lambda'_{ij})
%  \\
%  p(A_{jt}\given V_t) &=& \normal(A_{jt}\given 0,V_t)
%  \\
%  p(\mathrm{noise}_{ij}) &=& \normal(\mathrm{noise}_{ij}\given 0,\sigma_{ij}^2)
%  \quad ,
%\end{eqnarray}
%where the $A_{jt}$ are systematic noise amplitudes drawn from a
%Gaussians of variance $V_t$, the $G_{t}()$ are systematics functions
%(eigenfunctions, perhaps), and the pixel noise is all drawn
%independently (but heteroskedastically).
Note that nothing about this model requires that the wavelength grid
$\lambda_i$ be the same from spectrum to spectrum (that is, the
wavelength grid can depend on $j$), and there is no requirement that
the data at different epochs be similar in signal-to-noise. We do,
effectively, at this point, assume that instrument resolution is
identical (or very similar) from epoch to epoch.

The free parameters to be optimized are the scaling factors $s_j$, the basis expansion
coefficients $a_m$, and the radial velocities $\beta_j$. We use the $I\times J\times N_{par}$ matrix 
of derivatives of $y_{ij}$ in the optimization:
\begin{eqnarray}
   %\frac{dy_{ij}}{d \mathrm{par}} &=& \Big((\frac{\partial y_{ij}}{\partial s_j})^2 + 
   %\sum_{m=0}^{M-1}(\frac{\partial y_{ij}}{\partial f} \frac{\partial f}{\partial a_m})^2 + 
  %(\frac{\partial y_{ij}}{\partial f} \frac{\partial f}{\partial \beta})^2\Big)^{\frac{1}{2}}
  %\\
  \frac{\partial y_{ij}}{\partial s_k} &=& f(\lambda'_{ij}) \delta_{jk}
  \\
  \frac{\partial y_{ij}}{\partial f} \frac{\partial f}{\partial a_m} &=& s_j g_m(\lambda'_{ij})
  \\
  \frac{\partial y_{ij}}{\partial f} \frac{\partial f}{\partial \beta_k} &=& 
  %s_j \sum_{m=0}^{M-1} a_m \frac{\partial g_m}{\partial \lambda'_{ij}} 
  %\frac{\partial \lambda'_{ij}}{\partial \beta_k}  \delta_{jk} = 
  - s_j \sum_{m=0}^{M-1} a_m g'_m(\lambda'_{ij}) \frac{\lambda'_{ij}}{(1 - \beta_k)(1 + \beta_k)}
  \delta_{jk}
  \quad ,
\end{eqnarray}
where $\delta$ is the Dirac delta function and $g'(\lambda)$ is the first derivative of equation \ref{eqn:triangle},
defined as:
\begin{equation}
  g_m(\lambda) = \left\{
        \begin{array}{ll}
            0 & \quad \lambda \leq x_m -\Delta x \\
            \frac{1}{\Delta x} & \quad x_m - \Delta x \leq \lambda \leq x_m \\
            - \frac{1}{\Delta x} & \quad x_m \leq \lambda \leq x_m + \Delta x \\
            0 & \quad \lambda \geq x_m + \Delta x
        \end{array}
    \right.
    \quad .
\end{equation}

\paragraph{Acknowledgements:}
Thanks to DFM for helping with the model.

\end{document}
